Division and Factorization of p-Adic Polynomials

### 2016-01-03 polynomials/muldivfact.tex

@@

@@

\title{Multiplication, division and factorization\\of p-adic polynomials}

@@

\begin{deftn}
A \emph{Newton function of degree $n$} is a convex function 
$\varphi : [0,n] \to \R$ which is piecewise affine and whose epigraph
$\Epi(\varphi)$ have integral extremal points.
\end{deftn}

@@

We recall that one can attach a Newton function to each polynomial.
If $P(X) = \sum_{i=0}^n a_n X^n \in K_n[X]$, we define its Newton
polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
($1 \leq i \leq n$) together with the point at infinity $(0,+\infty)$
and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
function whose epigraph is $\NP(P)$. It is well known \todo{(Give a
reference?)} that:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NP(P+Q) & \subset & \text{Conv}\big(\NP(P) \cup \NP(Q)\big) \smallskip \\
\NP(PQ) & = & \NP(P) + \NP(Q)
\end{array}$$
where $\text{Conv}$ denotes the convex hull and the plus sign stands 
for the Minkowski sum. This translates to:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NF(P+Q) \geq \NF(P) \oplus \NF(Q) \smallskip \\
\NF(PQ) = \NF(P) \odot \NF(Q)
\end{array}$$
where the operations $\oplus$ and $\odot$ are defined accordingly.
It turns out that there exist efficient algorithms for computing these 
operations.

@@

It is easily checked that if $P$ and $Q$ are two polynomials known at 
precision $O(\varphi_P)$ and $O(\varphi_Q)$ respectively, then $P+Q$ is 
known at precision $O(\varphi_P \oplus \varphi_Q)$ and $PQ$ is known at 
precision $O\big(\varphi_P \odot \NF(Q) \oplus \NF(P) \odot 
\varphi_Q\big)$.

@@

As already mentioned, the jagged precision model is a particular case
of the lattice precision. Indeed, a precision of the shape
$\sum_{i=0}^n O(\pi^{N_i}) X^i$ corresponds to the lattice
generated by the elements $\pi^{N_i} X^i$ ($0 \leq i \leq n$).

@@

\begin{theo} \label{theo:EDivisionNP}
Here should be the theorem for the Newton polygons during a Euclidean division.
\end{theo}

\section{Modular multiplication}

\section{Slope factorization}
\subsection{Main theorem}

\begin{theo} \label{theo:slope-factor}
Under some assumptions... $\textrm{coeff}(P,d)=1,$ $\NP (A_0)' <0$ (???)

\begin{align*}
A_0 &= \textrm{trunc}(P,d), \\
B_{0} &= P \div A_{0}, \\
V_{0} &= 1.
\end{align*}

\begin{align*}
A_{i+1} &= A_i + (V_i P \% A_i), \\
B_{i+1} &= P \div A_{i+1}, \\
V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \% A_{i+1}.
\end{align*}
$A_i$ convergences to $A_\infty$ such that $\NP (A_\infty)= \NP (A_0)$ and $A_\infty$ divides $P.$
\end{theo}

\begin{rem}
Using $P \mapsto P^*=X^{deg(P)}P \left( \frac{1}{X} \right), $ the assumptions in Theorem \ref{theo:slope-factor} do not induce any loss in generality.
\end{rem}

\begin{rem}
Non necessity to have $A_0$ as the first slope. It is possible to use minorations.

\draw (0,1)-- (1.5,0) -- (3,-.375)-- (3.5,.625);
\draw[green] (0,1.5)-- (1,0.33)-- (1.5,0) --(2,-.125)-- (3,0)-- (3.5,1);

\node[below] at (.75,-0.2) {$\lambda_0$};
\end{tikzpicture}
\end{rem}

\subsection{Useful formulae}

\begin{lem} \label{lem:formulae}
Some definitions :
\begin{align}
R_i &= A_{i+1} -A_i, \label{eqdef:Ri}\\
S_i &= P \% A_i, \label{eqdef:Si} \\
Q_i &= V_i P \div A_i. \label{eqdef:Qi}. 
\end{align}
Some formulae :
\begin{align}
B_i-B_{i+1} &= R_i B_{i+1} \div A_i, \label{eqdef:Biminus} \\
S_i-S_{i+1} &= R_i B_{i+1} \div A_i,  \label{eqdef:Siminus} \\
S_{i} &= (B_i R_i + (1-V_i B_i) S_{i-1}+(1-V_i B_i)(S_i-S_{i-1})) \% A_{i}, \label{eqdef:Si2}  \\
V_{i+1}-V_i &= (V_i (1-V_i B_i)+V_i^2(B_i-B_{i+1})) \% A_{i}, \label{eqdef:Viminus}\\
1-Q_i &= (1-V_i B_i)+((R_i-V_i S_i) \div A_i), \label{eqdef:Qiminus}\\
R_{i+1} &= ((V_{i+1}-V_i)S_{i+1}+(1-Q_i)R_i) \% A_{i+1}, \label{eq:Riplus}\\
1-V_{i+1}B_{i+1} &= ((1-V_i B_i)+V_i (B_i-B_{i+1}))^2. \label{eq:ViBiplus}
\end{align}
\end{lem}
\begin{proof}
We first remark that by definition,
\begin{align*}
P &= A_i B_i + S_i, \\
P V_i&= A_i Q_i + R_i.
\end{align*}
Since $P= A_i B_i + S_i= A_{i+1} B_{i+1} + S_{i+1},$ we have \[R_i B_{i+1}= (B_i -B_{i+1})A_i + (S_i+S_{i+1}).\]
Hence, by consideration of degree, what we have written is the Euclidean division of $R_i B_{i+1}$ by $A_i$ and we obtain \eqref{eqdef:Biminus} and \eqref{eqdef:Siminus}.

From $V_i P =  A_i Q_i + R_i=V_i (A_i B_i + S_i),$ we get \begin{equation}(V_i B_i -Q_i) A_i=R_i-V_i S_i. \label{eq:ViBiQi} \end{equation} Thus, \[ R_i=V_i S_i \% A_i.\]
Hence, $B_i R_i =S_i (V_iB_i-1)+S_i \% A_i,$ and $S_i=B_i R_i-S_i (V_iB_i-1) \% A_i.$ \eqref{eqdef:Si2} follows directly.

By definition of $V_i,$ we get $V_{i+1}-V_i=V_i (1-V_i B_{i+1}) \% A_{i+1}.$ \eqref{eqdef:Viminus} follows immediately.

We can write $1-Q_i=(1-V_i B_i)+(V_i B_i - Q_i).$ Using \eqref{eq:ViBiQi}, we get \eqref{eqdef:Qiminus}.

We have $V_i P=A_i Q_i+R_i=(A_{i+1}-R_i)Q_i+R_i,$ and $V_{i+1} P=A_{i+1} Q_{i+1}+R_{i+1}.$ As a consequence, 
\begin{align*}
R_{i+1} &= (V_{i+1}-V_i)P+(1-Q_i)R_i \\
 &= (V_{i+1}-V_i)S_{i+1}+(1-Q_i)R_i) \% A_{i+1},
\end{align*} and \eqref{eq:Riplus} is proved.

Finally, \begin{align*}
1-V_{i+1}B_{i+1} &= 1-2V_i B_{i+1}+V_i^2 B_{i+1}^2 \% A_{i+1}, \\
&= (1-V_i B_{i+1})^2 \% A_{i+1}, \\
&= ((1-V_i B_i)+V_i (B_i-B_{i+1}))^2 \% A_{i+1},
\end{align*} which concludes the proof.
\end{proof}

>>>>

\renewcommand{\mod}{\,\%\,}
\renewcommand{\div}{\,/\hspace{-0.3em}/\,}
\newcommand{\nfop}[1]{%
\raisebox{-0.1mm}{\begin{tikzpicture}
\draw[transparent] (-0.18,0)--(0.18,0);
\draw[very thin] (0,0) circle (0.11cm);
\begin{scope}
\clip (0,0) circle (0.11cm);
\node[scale=0.75] at (0,0) { $#1$ };
\end{scope}
\end{tikzpicture}}}
\newcommand{\nfplus}{\nfop+}
\newcommand{\nftimes}{\nfop\times}
\newcommand{\nfmod}{\nfop\mod}
\newcommand{\nfdiv}{\nfop\div}

@@

\newcommand{\tinyplus}{\raisebox{0.4mm}{\hspace{0.4mm}\tiny +\hspace{0.4mm}}}

@@

\title{Division and Factorization of p-Adic Polynomials}

@@

\begin{deftn}
A \emph{Newton function of degree $n$} is a convex function 
$\varphi : [0,n] \to \R \cup \{+\infty\}$ which is piecewise affine, 
which takes a finite value at $n$ and whose epigraph $\Epi(\varphi)$ 
have integral extremal points.
\end{deftn}

@@

We recall that one can attach a Newton function to each polynomial.
If $P(X) = \sum_{i=0}^n a_n X^n \in K_n[X]$, we define its Newton
polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
($1 \leq i \leq n$) together with the point at infinity $(0,+\infty)$
and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
function whose epigraph is $\NP(P)$. It is well known \todo{(Give a
reference?)} that:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NP(P+Q) & \subset & \text{Conv}\big(\NP(P) \cup \NP(Q)\big) \smallskip \\
\NP(PQ) & = & \NP(P) + \NP(Q)
\end{array}$$
where $\text{Conv}$ denotes the convex hull and the plus sign stands 
for the Minkowski sum. This translates to:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NF(P+Q) \geq \NF(P) \nfplus \NF(Q) \smallskip \\
\NF(PQ) = \NF(P) \nftimes \NF(Q)
\end{array}$$
where the operations $\nfplus$ and $\nftimes$ are defined accordingly.
There exist classical algorithms for computing these two operations
whose complexity is quasi-linear with respect to the degree.

@@

It is easily checked that if $P$ and $Q$ are two polynomials known at 
precision $O(\varphi_P)$ and $O(\varphi_Q)$ respectively, then $P+Q$ is 
known at precision $O(\varphi_P \nfplus \varphi_Q)$ and $PQ$ is known at 
precision $O\big((\varphi_P \nftimes \NF(Q)) \nfplus (\NF(P) \nftimes 
\varphi_Q)\big)$.

\begin{deftn}
\label{def:nondeg}
Let $P = P_0 + O(\varphi_P)$. We say that the Newton precision 
$O(\varphi_P)$ on $P$ is \emph{nondegenerate} if $\varphi_P \geq 
\NF(P_0)$ and $\varphi_P(x) > y$ for all extremal point $(x,y)$ of 
$\NP(P_0)$.
\end{deftn}

We notice that, under the conditions of the above definition, the
Newton polygon of $P$ is well defined. Indeed, if $\delta P$ is any
polynomial whose Newton function is not less than $\varphi_P$, we
have $\NP(P_0 + \delta P) = \NP(P_0)$.

@@

As already mentioned, the jagged precision model is a particular case of 
the lattice precision. Indeed, a precision of the shape $\sum_{i=0}^n 
O(\pi^{N_i}) X^i$ corresponds to the lattice generated by the elements 
$\pi^{N_i} X^i$ ($0 \leq i \leq n$). This remark is the origin of the 
notion of \emph{diffused digits of precision} introduced in 
\cite[Definition 2.3]{preclinalg}. We shall use it repeatedly in the 
sequel in order to compare the behaviour of the three aforementioned 
precision data in concrete situations.

@@

\todo{Add introduction.}
In the sequel, we shall will the notation $A \div B$ (resp $A \mod B$)
for the quotient (resp. the remainder) in the Euclidean division of $A$ 
by $B$.

\subsection{Euclidean division of Newton functions}

\begin{deftn}
Let $\varphi$ and $\psi$ be two Newton functions of degree $n$ and
$d$ respectively. Letting $\mu = \psi(d) - \psi(d-1)$ and $\Delta$
be the greatest affine function of slope $\mu$ with $\Delta \leq
\varphi$, we define:
$$\begin{array}{rcl}
\varphi \nfmod \psi \, : \, [0, d-1] & \to & \R \cup \{+\infty\} \smallskip \\
x & \mapsto & \psi(x) + \Delta(d) - \psi(d) \medskip \\
\varphi \nfdiv \psi \, : \, [0, n-d] & \to & \R \cup \{+\infty\} \smallskip \\
x & \mapsto & \inf_{h \geq 0} \psi(x+d+h) - \mu h.
\end{array}$$
\end{deftn}

\begin{figure}
\hfill
\begin{tikzpicture}[xscale=0.7, yscale=0.5]
\fill[blue!30] (4,9.5)--(4,2.667)--(2,2)--(0.5,4)--(0,8)--(0,9.5)--cycle;
\fill[yellow!50] (5,9.5)--(5,3)--(4,2.667)--(4,9.5)--cycle;
\fill[green!30] (5,9.5)--(5,3)--(8,4)--(11,8)--(11,9.5)--cycle;
\draw[->] (-0.5,0)--(11.5,0);
\draw[->] (0,-0.5)--(0,9.5);
\draw[very thick] (0,6)--(0.5,2)--(2,0)--(5,1);
\draw[very thick] (0,9)--(3,5)--(8,4)--(11,8);
\draw[dashed] (11,5)--(8,4);
\draw (8,4)--(2,2)--(0.5,4)--(0,8);
\draw[dotted,thick] (5,0)--(5,9.5);
\draw[dotted,thick] (4,0)--(4,9.5);
\node[scale=0.8, below left] at (0,0) { $0$ };
\node[scale=0.8, below] at (5,0) { $d$ };
\node[scale=0.8, below] at (4,0) { $d{-}1$ };
\node[scale=0.9, below right] at (2.5,1.2) { $\psi$ };
\node[scale=0.9, right] at (9.8,7.3) { $\varphi$ };
\node[scale=0.9, right] at (10.5,4.4) { $\Delta$ };
\end{tikzpicture}
\hfill \null

\vspace{-6mm}

\caption{Euclidean division of Newton functions}
\label{fig:NewtonEuclide}
\end{figure}

Figure~\ref{fig:NewtonEuclide} illustrates the definition: if $\varphi$ 
and $\psi$ are the functions represented on the diagram, the epigraph of 
$\varphi \nfmod \psi$ is the blue area whereas that of $\varphi \nfdiv \psi$ 
is the green area translated by $(-d,0)$.
It is an easy exercise (left to the reader) to design algorithms for
computing $\varphi \nfmod \psi$ and $\varphi \nfdiv \psi$ for a cost
which is at most $n + O(1)$.

\begin{theo}
\label{theo:EDivisionNP}
Let $A, B \in K[X]$ with $B \neq 0$. Let $d$ be the degree of $B$.
Let $\mu = \NF(B)(d) - \NF(B)(d{-}1)$ be the last slope of $B$ and 
$y = \Delta(x)$ be the
equation of the highest line of slope $\mu$ lying below $\NP(A)$.
Then:
\begin{align}
\NF(A \mod B) & \geq \NF(B) \nfmod \NF(A) \label{eq:AmodB} \\
\text{and} \hspace{4mm}
\NF(A \div B) & \geq \NF(B) \nfdiv \NF(A) \label{eq:AdivB}
\end{align}
\end{theo}

\begin{rem}
Keeping the notations of the Theorem and decomposing $A$ as a sum 
$A_{<d} + A_{\geq d}$ where $A_{<d}$ (resp. $A_{\geq d}$) consists of
monomials of $A$ of degree $< d$ (resp of degree $\geq d$), we have:
$$A \div B = A_{\geq d} \div B 
\quad \text{and} \quad
A \mod B = A_{<d} + (A_{\geq d} \mod B).$$
Using the above relations and applying Theorem~\ref{theo:EDivisionNP}
with $A_{\geq d}$ and $B$ (instead of $A$ and $B$ respectively) may lead
to a better lower bound on the Newton function of $A \div B$ and $A \mod
B$.
\end{rem}

\begin{proof}[of Theorem~\ref{theo:EDivisionNP}]
Let us first prove Eq.~\eqref{eq:AmodB}. 
Replacing $B$ by $c^{-1} B$ where $c$ denotes the leading coefficient
of $B$, we may assume that $B$ is monic. Using linearity, we may further
assume that $A$ is a monomial. Set $R_n = X^n \mod B$. The relation
we have to prove is:
$$\NF(R_n)(x) \geq \NF(B)(x) - \mu(n-d).$$
When $n < d$, we have $R_n = 0$ and the assertion of the Theorem holds. 
For larger $n$, we proceed by induction. We have the relation
$R_{n+1} = X R_n - c_n B$
where $c_n$ is the coefficient in $X^{d-1}$ of $R_n$. Thanks to the
induction hypothesis, we have:
\begin{align*}
\val(c_n) \geq \NF(R_n)(d{-}1) & \geq \NF(B)(d{-}1) - \mu(n{-}d) \\
& = - \mu(n{+}1{-}d)
\end{align*}
since $\mu = -\NF(B)(d{-}1)$ because $B$ is monic. Therefore
$\NF(c_n B)(x) \geq \NF(B)(x) - \mu(n+1-d)$ for all $x$. On the 
other hand, for all $x$, we have:
$$\NF(X R_n)(x) = \NF(R_n)(x{-}1) \geq \NF(R_n)(x) - \mu$$
from what we get $\NF(X R_n)(x) \geq \NF(B)(x) - \mu(n+1-d)$. As a
consequence
$\NF(R_{n+1})(x) \geq \NF(B)(x) - \mu(n+1-d)$
and the induction follows.
Eq.~\eqref{eq:AdivB} is now derived from:
$$\NF(A \div B) \nftimes \NF(B) \geq \NF(A) \nfplus \NF(A \mod B)$$
using the estimation on $\NF(A\mod B)$ we have just proved
(see Figure~\ref{fig:NewtonEuclide}).
\end{proof}

\subsection{Tracking precision}

\noindent
{\bf Jagged precision.}
\todo{Write something... Explain in particular that the fast methods
(based on fast multipliction) may lead to important looses of precision.}

\noindent
{\bf Newton precision.}
We consider here the case where the precision is given by Newton
functions. Concretely, we pick $A, B \in K[X]$ two polynomials which 
are known at precision $O(\varphi_A)$ and $O(\varphi_B)$ respectively:
$$A = A_0 + O(\varphi_A)
\quad \text{and} \quad
B = B_0 + O(\varphi_B)$$
Here $A_0$ and $B_0$ are some approximations of $A$ and $B$ respectively 
and $\varphi_A$ and $\varphi_B$ denotes two Newton functions of degree 
$\deg A$ and $\deg B$ respectively.
We are interesting in determining the precision on $A \mod B$ and $A
\div B$. The following proposition gives a theoretical answer under
mild assumptions.

\begin{prop}
\label{prop:NewtonprecEuclide}
We keep the above notations and assume that the Newton precisions
$O(\varphi_A)$ and $O(\varphi_B)$ on $A$ and $B$ respectively are 
both nondegenerate (\emph{cf} Definition~\ref{def:nondeg}). Then,
setting:
$$\varphi = \varphi_A \nfplus 
\big[ \varphi_B \nftimes \big(\NF(A) \nfdiv \NF(B)\big) \big]$$
the polynomials $A \div B$ and $A \mod B$ are known at precision 
$O(\varphi \nfdiv \NF(B))$ and $O(\varphi \nfmod \NF(B))$ respectively.
\end{prop}

\begin{proof}
Let $\delta A$ (resp. $\delta B$) be a polynomial whose Newton 
function is not less than $\varphi_A$ (resp. $\varphi_B$) and define
$\delta Q$ and $\delta R$ by:
\begin{align*}
Q_0 + \delta Q & = (A_0 + \delta A) \div (B_0 + \delta B) \\
R_0 + \delta R & = (A_0 + \delta A) \mod (B_0 + \delta B)
\end{align*}
where $Q_0 = A_0 \div B_0$ and $R_0 = A_0 \mod B_0$.
We have to show that $\NF(\delta Q) \geq \varphi \nfdiv \NF(B)$ and
$\NF(\delta R) \geq \varphi \nfmod \NF(B)$.
Set $\delta X = \delta A - Q_0 \delta B$. 
Using Theorem~\ref{theo:EDivisionNP}, we obtain $\NF(Q_0) \geq
\NF(A) \div \NF(B)$ and consequently $\NF(\delta X) \geq \varphi$.
On the other hand, an easy computation yields
$$\delta X = (B + \delta B) \cdot \delta Q + \delta R$$
so that $\delta Q = \delta X \div (B + \delta B)$ and 
$\delta R = \delta X \mod (B + \delta B)$. Using again 
Theorem~\ref{theo:EDivisionNP}, we get the desired result.
\end{proof}

\todo{Explain how one can use this proposition: we split the computation
into two parts (approximation and precision)...\\
Don't forget to give the complexity.}

\noindent
{\bf Lattice precision.}
We now move to lattice precision. We pick $A$ and $B$ two polynomials 
of respective degree $n$ and $d$ and assume that they are known at 
precision $O(H_A)$ and $O(H_B)$ respectively:
$$A = A_0 + O(H_A)
\quad \text{and} \quad
B = B_0 + O(H_B).$$
where $H_A \in K_{\leq n}[X]$ and $H_B \in K_{\leq d}[X]$ are lattices. 
According to the results of \cite{padicprec}, in order to determine the 
precision on $A \div B$ and $A \mod B$, we need to compute the 
differential of the mappings $(X,Y) \mapsto X \div Y$ and $(X,Y)
\mapsto X \mod Y$ at the point $(A_0, B_0)$. Writing $Q_0 = A_0 \div
B_0$ and $R_0 = A_0 \mod B_0$, this can be done by
expanding the relation:
$$A_0 + dA = (B_0 + dB) (Q_0 + dQ) + (R_0 + dR)$$
and neglecting the terms of order $\geq 2$. We get this way 
$dA = B_0 dQ + Q_0 dB + dR$
meaning that $dQ$ and $dR$ appears respectively as the quotient and
the remainder of the Euclidean division of $dX = dA - B_0 dQ$ by $B_0$.

Once this has been done, the strategy is quite similar to that explained 
for Newton precision: \todo{summerize this in one sentence.}

\subsection{An example: modular multiplication}

For this example, we work over $W = \Z_2$ and fix a monic polynomial $M 
\in \Z[X]$ (known exactly) of degree $5$. Our aim is to compare the 
numerical stability of the multiplication in the quotient $\Z_2[X]/M(X)$ 
depending on the precision model we are using. In order to do so, we 
pick $n$ random polynomials $P_1, \ldots, P_n$ in $\Z_2[X]/M(X)$ 
(according to the Haar measure) whose coefficients are all known at 
precision $O(2^N)$ for some large integer $N$. We then compute the 
product of the $P_i$'s using the following quite naive algorithm.

\noindent\hrulefill

%\noindent {\bf Algorithm 1:} {\tt example\_product}
%
%\smallskip

\noindent 1.\ {\bf set} $P = 1$

\noindent 2.\ {\bf for} $i=1,\dots,n$ {\bf do} {\bf compute} $P = (P 
\cdot P_i) \mod M$

\noindent 3.\ {\bf return} $P$

\vspace{-1ex}\noindent\hrulefill

The table below reports the average gain of \emph{absolute} 
precision $G$ which is observed while executing the algorithm above 
for various modulus and $n$. The average is taken on a sample of $1000$ 
random inputs. We recall that $G$ is defined as follows:

\noindent $\bullet$
in the case of jagged and Newton precision, the precision on the output 
may be written into the form $\sum_{i=1}^4 O(2^{N_i}) \: X^i$ and 
$G = \sum_{i=1}^4 (N_i - N)$;

\noindent $\bullet$
in the case of lattice precision, the precision on the output is a 
lattice $H$ and $G$ is the index of $H$ in $2^N \mathcal L$ where 
$\mathcal L = \Z_2[X]/M$ is the standard lattice; in that case, we write 
$G$ as a sum $G_{\text{nd}} + G_{\text{d}}$ where $G_{\text{d}}$ is the 
index of $H$ in the largest lattice $H_0$ contained in $H$ which can be 
generated by elements of the shape $2^{N_i} X^i$ ($0 \leq i \leq 4$). 
(The term $G_d$ corresponds to diffused digits according to 
\cite[Definition 2.3]{preclinalg}.)

{\small%
\noindent \hfill%
\renewcommand{\arraystretch}{1.2}%
\begin{tabular}{|c|c|c|c|c|}%
\hline
\multirow{3}{*}{Modulus $M$} & 
\multirow{3}{*}{\hspace{2mm}$n$\hspace{2mm}} & 
\multicolumn{3}{c|}{Gain of precision} \\
\cline{3-5}
& & \multirow{2}{*}{Jagged} & \multirow{2}{*}{Newton}
      & \raisebox{-0.5mm}{Lattice} \\
& & & & \raisebox{0.5mm}{\scriptsize (not dif.\tinyplus dif.)} \\
\hline
\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + X^2 + 1$ \\ {\scriptsize (Irred. mod $2$)} \end{tabular}}
& $10$ & $\phantom{00}0.2$ & $\phantom{00}0.2$ & $\phantom{00}0.2\tinyplus\phantom{00}0.0$ \\
& $50$ & $\phantom{00}4.2$ & $\phantom{00}4.2$ & $\phantom{00}4.2\tinyplus\phantom{00}0.0$ \\
& $100$ & $\phantom{0}11.2$ & $\phantom{0}11.2$ & $\phantom{0}11.2\tinyplus\phantom{00}0.0$ \\
\hline
\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + 1$ \\ {\scriptsize (Sep. mod $2$)} \end{tabular}}
& $10$ & $\phantom{00}0.4$ & $\phantom{00}0.4$ & $\phantom{00}0.9\tinyplus\phantom{00}6.0$ \\
& $50$ & $\phantom{00}5.6$ & $\phantom{00}5.6$ & $\phantom{0}11.1\tinyplus\phantom{0}42.0$ \\
& $100$ & $\phantom{0}13.6$ & $\phantom{0}13.6$ & $\phantom{0}27.0\tinyplus\phantom{0}87.0$ \\
\hline
\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + 2$ \\ {\scriptsize (Eisenstein)} \end{tabular}}
& $10$ & $\phantom{00}6.2$ & $\phantom{00}6.2$ & $\phantom{00}6.2\tinyplus\phantom{00}0.0$ \\
& $50$ & $\phantom{0}44.0$ & $\phantom{0}44.0$ & $\phantom{0}44.0\tinyplus\phantom{00}0.0$ \\
& $100$ & $\phantom{0}92.5$ & $\phantom{0}92.5$ & $\phantom{0}92.5\tinyplus\phantom{00}0.0$ \\
\hline
\multirow{3}{*}{\begin{tabular}{@{}c@{}} $(X+1)^5 + 2$ \\ {\scriptsize (Shift Eisenstein)} \end{tabular}}
& $10$ & $\phantom{00}0.6$ & $\phantom{00}0.6$ & $\phantom{00}4.7\tinyplus\phantom{00}1.4$ \\
& $50$ & $\phantom{00}7.1$ & $\phantom{00}7.1$ & $\phantom{0}42.6\tinyplus\phantom{00}1.4$ \\
& $100$ & $\phantom{0}15.1$ & $\phantom{0}15.1$ & $\phantom{0}91.8\tinyplus\phantom{00}1.4$ \\
\hline
\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + X + 2$ \\ {\scriptsize (Two slopes)} \end{tabular}}
& $10$ & $\phantom{00}1.7$ & $\phantom{00}1.7$ & $\phantom{00}7.9\tinyplus\phantom{00}9.8$ \\
& $50$ & $\phantom{00}8.1$ & $\phantom{00}8.1$ & $\phantom{0}70.7\tinyplus\phantom{0}59.8$ \\
& $100$ & $\phantom{0}16.1$ & $\phantom{0}16.1$ & $152.6\tinyplus125.9$ \\
\hline
\end{tabular}
\hfill \null}

We observe several interesting properties. First of all, the gains for 
Newton precision and jagged precision always agree though one may has 
thought at first that Newton precision is weaker. Since performing
precision computations in the Newton framework is cheaper, it seems
(at least on this example) that using the jagged precision model is not
relevant.

On the other hand, the lattice precision may end up with better results. 
Nevertheless this strongly depends on the modulus $M$. For instance, 
when $M$ is irreducible modulo $p=2$ or Eiseistein, there is apparently 
no benefit of using the lattice precision model. We emphasize that 
these two particular cases correspond to modulus that are usually used 
to define (unramified and totally ramified respectively) extensions of 
$\Q_2$.

For other modulus, the situation is quite different and the benefit of 
using the lattice precision model becomes more apparent. 
The comparison between the gain of precision in the jagged model and 
the number of not diffused digits in the lattice model makes sense:
indeed the latter appears as a theoretical upper bound of the former
and the difference between them quantifies the quality of the way we
track precision in the jagged (or the Newton) precision model. We 
observe that this difference is usually not neglictible (\emph{cf} 
notably the case of $M(X) = (X+1)^5 + 2$) meaning that this quality 
is not very good in general.
As for diffused digits, they correspond to digits that cannot be 
``seen'' in the jagged precision model. Their number then measures the 
intrinsic limitations of this model. We observe that it can be very 
important as well in several cases.

\todo{Add some words about the comparison between $X^5 + 2$ and
$(X+1)^5 + 2$?}

\section{Slope factorization}

A well-known theorem \todo{(Give a reference?)} asserts that each 
extremal point $M$ in the Newton polygon $\NP(P)$ of a polynomial $P \in 
K[X]$ corresponds to a factorization $P = AB$ where the Newton polygon 
of $A$ (resp. $B$) is given by the part of $\NP(P)$ located at the left 
(resp. the right) of $M$. Such a factorization is often called a 
\emph{slope factorization}.

The aim of this section is to design efficient and stable algorithms 
for computing these factorizations. \todo{Compare with previous
results...}

\subsection{A Newton iteration}

In the previous references, the factor $A$ defined above is always 
obtained \emph{via} a Newton iteration which takes place over a finite 
extension of $K$. We introduce here a variant of this Newton iteration 
whose main advantage to be defined over $K$ itself.

\begin{theo} \label{theo:slope-factor}
Let $P(X) = \sum_{i=0}^n a_i X^i$ be a polynomial of degree $n$ with
coefficients in $K$. We assume that $\NP(P)$ has an extremal point 
whose abscissa is $d$.
We define the sequences $(A_i)_{i \geq 0}$ and $(V_i)_{i \geq 0}$
recursively by:
\begin{align*}
A_0 & = \sum_{i=0}^d a_i X^i, \quad V_{0} = 1 \medskip \\
A_{i+1} &= A_i + (V_i P \:\mod\: A_i), \smallskip \\
V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \mod A_{i+1} \\
& \hspace{2cm}\text{where } B_{i+1} = P \div A_{i+1}.
\end{align*}
Then the sequence $(A_i)$ convergences to a divisor $A$ of $P$ 
of degree $d$ whose leading coefficient is $a_d$ and whose Newton 
function agrees with $\NF(P)$ on $[0,d]$.
\end{theo}

%\begin{rem}
%Using $P \mapsto P^*=X^{deg(P)}P \left( \frac{1}{X} \right), $ the assumptions in Theorem \ref{theo:slope-factor} do not induce any loss in generality.
%\end{rem}

%\begin{rem}
%Non necessity to have $A_0$ as the first slope. It is possible to use minorations.
%
%\begin{tikzpicture}
%\draw[->] (0,-1) -- (0,2);
%\draw[->] (-0.05,0) -- (4,0);
%\node[left] at (0,0) {$0$};
%\draw (1.5,-.1) -- (1.5,.1);
%\node[above] at (1.5,0) {$d$};
%\draw (0,1)-- (1.5,0) -- (3,-.375)-- (3.5,.625);
%\draw[green] (0,1.5)-- (1,0.33)-- (1.5,0) --(2,-.125)-- (3,0)-- (3.5,1);
%\node[below] at (.75,-0.2) {$\lambda_0$};
%\end{tikzpicture}
%\end{rem}

\begin{rem}
The divisor $A$ is uniquely determined by the conditions of 
Theorem~\ref{theo:slope-factor}.
\end{rem}

The rest of this subsection is devoted to the proof of the theorem.
If $d = n$ (resp. $d = 1$), the sequence $A_i$ is constant equal to
$P$ (resp. to the constant coefficient of $P$) and theorem is clear.
We then assume $1 < d < n$. Without loss of generality, we may further 
assume that the coefficient $a_d$ is equal to $1$. We set:
\begin{align*}
\lambda_0 & = \NF(P)(d) - \NF(P)(d{-}1) \\
\lambda_1 & = \NF(P)(d{+}1) - \NF(P)(d).
\end{align*}
The existence of an extremal point of $\NP(P)$ located at abscissa
$d$ ensures that $\lambda_1 > \lambda_0$.
For all indices $i$, we define:
$$\begin{array}{r@{\,\,}lr@{\,\,}l}
Q_i &= V_i P \div A_i, &
R_i & = V_i P \mod A_i = A_{i+1} - A_i, 
\smallskip \\
S_i &= P \mod A_i, &
T_i &= 1 - V_i B_i. 
\end{array}$$
and when $\square$ is some letter, we put $\Delta \square_i = \square_{i+1} 
- \square_i$.
%We use the next Lemma to record some useful formulae.

\begin{lem} \label{lem:formulae}
The following relations hold:
\begin{align}
\Delta B_i &= -(R_i B_{i+1}) \div A_i, \label{eqdef:Biminus} \\
\Delta S_i &= -(R_i B_{i+1}) \mod A_i,  \label{eqdef:Siminus} \\
S_{i} &= (B_i R_i + T_i S_{i-1} - T_i \: \Delta S_{i-1}) \mod A_{i}, \label{eqdef:Si2}  \\
\Delta V_i &= (V_i T_i - V_i^2 \: \Delta B_i) \mod A_{i}, \label{eqdef:Viminus}\\
1-Q_i &= T_i + (R_i-V_i S_i) \div A_i, \label{eqdef:Qiminus}\\
R_{i+1} &= (\Delta V_i \: S_{i+1}+(1{-}Q_i)R_i) \mod A_{i+1}, \label{eq:Riplus}\\
T_{i+1} &\equiv T_i + V_i \: (\Delta B_i)^2 \pmod{A_{i+1}}. \label{eq:ViBiplus}
\end{align}
\end{lem}

\begin{proof}
From $P= A_i B_i + S_i= A_{i+1} B_{i+1} + S_{i+1}$, we get
$- R_i B_{i+1}= \Delta B_i \cdot A_i + \Delta S_i$.
Hence, by consideration of degree, we obtain \eqref{eqdef:Biminus} 
and \eqref{eqdef:Siminus}.
On the other hand, from $V_i P = A_i Q_i + R_i=V_i (A_i B_i + S_i)$, we 
derive
\begin{equation}
\label{eq:ViBiQi} 
(V_i B_i -Q_i) \cdot A_i=R_i-V_i S_i. 
\end{equation} 
Thus $R_i=V_i S_i \mod A_i$.
Hence $B_i R_i = (S_i - S_i T_i) \mod A_i$ and $S_i = B_i R_i + S_i T_i 
\mod A_i$, from what \eqref{eqdef:Si2} follows directly.
By definition of $V_i$, we get $\Delta V_i = V_i (1-V_i B_{i+1}) \mod A_{i+1}$ and consequently \eqref{eqdef:Viminus}.
We now write $1-Q_i= T_i + (V_i B_i - Q_i)$. Using \eqref{eq:ViBiQi}, we 
get \eqref{eqdef:Qiminus}.
